The Unicode Cookbook \linebreak for Linguists

### 2015-11-15 book/chapters/ipa_background.tex
readability update

Established in 1886, the International Phonetic Association (henceforth
Association) has long maintained a standard alphabet, the International Phonetic
Alphabet (IPA), which is a common standard in linguistics to transcribe sounds
of spoken languages. It was first published in 1888 as an international system
of phonetic transcription for oral languages and for pedagogical purposes. It
contained phonetic values for English, French and German. Diacritics for length
and nasalization were already present in this first version, and the same
symbols are still used today. 
%\footnote{Also referred to as API, for \textit{Association Phonétique Internationale}.}

Originally, the IPA alphabet was a list of symbols with pronunciation examples
from words in different languages. In 1900 the symbols were first organized into
a chart and were given phonetic feature labels for consonants (e.g.\ for manner
of articulation among others `plosives', `nasales', `fricatives' and for place
of articulation among others `bronchiales', `laryngales', `labiales') and for
vowels (e.g.\ `fermées', `mi-fermées', `mi-ouvertes', `ouvertes'). Throughout
the last century, the structure of the chart has changed with increases in
phonetic knowledge. Thus, similar to notational systems in other scientific
disciplines, the IPA reflects facts and theories of phonetic knowledge that have
developed over time. It is natural then that the IPA is modified occasionally to
accommodate scientific innovations and discoveries. In fact, this is part of the
Association's mandate. These changes are captured in the recurring revisions to
the IPA.
%\footnote{For a detailed history see:
%\url{https://en.wikipedia.org/wiki/History\_of\_the\_International\_Phonetic_Alphabet}}

Over the years there have been several revisions, many minor. For example, the
articulation labels -- what often are call `features' even though the IPA
deliberately avoids this term -- have changed (e.g.\ terms like `lips', `throat'
or `rolled' are not used anymore). Phonetic symbol values have changed (e.g.\
voiceless is no longer marked by <h>). Symbols have been dropped (e.g.\ the
caret diacritic denoting `long and narrow' is no longer used). And many symbols
have been added to reflect contrastive sounds found in the world's very diverse
phonological systems. The current construction and use of the IPA are guided by
principles outlined in the \textit{Handbook of the International Phonetic
Association: A guide to the use of the International Phonetic Alphabet}
\citep[159]{IPA1999}, henceforth simply called \textit{Handbook}.

@@

	\item it is intended to be a set of symbols for representing all possible
       sounds in the world's (spoken) languages
	\item its chart has a linguistic basis (specifically a phonological bias)
       rather than a general phonetic notation scheme
	\item its symbols can be used to represent distinctive feature
       combinations\footnote{Although the chart uses traditional manner and
       place of articulation labels, the symbols can nevertheless represent any
       defined bundle of features, binary or otherwise, to define phonetic
       dimensions.}
	\item its chart provides a summary of linguists' agreed-upon phonetic knowledge
       (a common denominator of phonological ``facts'')

@@

Several styles of transcription with IPA are possible, as illustrated in the
\textit{Handbook}, and they are all valid (see the 29 languages and their
transcriptions in the original and initial \textit{Illustrations of the IPA}
\citep[41--154]{IPA2007}). Therefore, there are different but equivalent
transcriptions, or as \citet[64]{Ladefoged1990a} captures it, ``Perhaps now that
the Association has been explicit in its eclectic approach, outsiders to the
Association will no longer speak of \textit{the} IPA transcription of a given
phenomenon, as if there were only one approved style.'' Clearly not all phoneticians agree (or will likely ever agree of course) on all 
aspects of the IPA or on transcription practices. As noted above, there have 
been several revisions in the IPA's long history, but the current version (2005) 
is strikingly similar to the 1926 version, showing the viability of the IPA as a 
common standard for linguistic transcription.

@@

Any IPA transcription is based on two premises: (i) that it is possible to
describe the acoustic speech signal (i.e.\ sound waves) in terms of sequentially
ordered discrete segments, and, (ii) that each segment can be characterized by
an articulatory target.\footnote{A purely phonetic description is only derivable
from instrumental data in high quality sound recordings. Once spoken language
data are segmented, phonological consideration inextricably play a roll in
transcription. In other words, phonetic observations beyond quantitative
acoustic analysis are always made in terms of some phonological framework.}
Today, the IPA chart reflects a linguistic theory grounded in principles of
phonological contrast and in knowledge about the attested linguistic variation.
This fact is stated explicitly in several places, including in the
\textit{Report on the 1989 Kiel convention} published in the \textit{Journal of
the International Phonetic Association} \citep[67-68]{International1989report}:

@@

Although the IPA provides symbols to unambiguously represent phonemes, it also
aims to represent phonetic details. Since phonetic detail could potentially
include anything, e.g. something like `deep voice', the IPA restricts phonetic
detail to linguistically relevant aspects of speech. From its earliest days the
Association aimed to provide ``a separate sign for each distinctive sound; that
is, for each sound which, being used instead of another, in the same language,
can change the meaning of a word'' \citep[27]{IPA1999}. Distinctive sounds
became later known as \textsc{phonemes} and the IPA has developed historically
into a notational devise with a strictly segmental phonemic view. A phoneme is
an abstract theoretical notion derived from an acoustic signal as produced by
speakers in the real world. Therefore the IPA contains a number of theoretical
assumptions about speech and how to analyze speech in written form.

@@

First, the symbols of IPA crucially distinguish between letter-like symbols and
diacritics.\footnote{see Section~\ref{pitfall-different-notions-of-diacritics} for a more
detailed discussion about the notion of diacritics} The use of the letter-like
symbols to represent a language's sounds is guided by the principle of contrast;
where two words are distinguishable by phonemic contrast, those contrasts should
be transcribed with different letter symbols (not diacritics). Allophonic
distinction falls under the rubric of diacritically-distinguished symbols, e.g.
the difference in English between and unrealized /p/ in [stop̚] vs. an aspirated
/p/ in [spʰot] should be marked with a diacritic. In sum:

@@

There are different reasons for this difficulty to interpret IPA symbols,
both officially part of the IPA specifications. Basically, the IPA allows for
both a `narrow' phonetic transcription and a `broad' phonological transcription.
An important principle of the IPA is that different representations resulting
from different underlying analyses are allowed. Because the IPA does not provide
phonological analyses for specific languages, the IPA does not define a single
``correct'' transcription system. Rather, the IPA aims to provide a resource
that allows users to express any analysis so that it is widely understood.

A `narrow' phonetic transcription may freely use the all symbols in the IPA
chart with direct reference to their phonetic value, i.e. the transcriber can
indicate with the symbols <ŋ͡m> that the phonetic value of the attested sound is
a simultaneous labial and velar closure which is voiced and contains nasal
airflow, independently of the phonemic status of this sound in the language in
question.

>>>>

Established in 1886, the \textit{International Phonetic Association} (henceforth
\textit{Association}) has long maintained a standard alphabet, \textit{the
International Phonetic Alphabet} (henceforth \textit{IPA}), which is a common
standard in linguistics to transcribe sounds of spoken languages. It was first
published in 1888 as an international system of phonetic transcription for oral
languages and for pedagogical purposes. It contained phonetic values for
English, French and German. Diacritics for length and nasalization were already
present in this first version, and the same symbols are still used today. 
%\footnote{Also referred to as API, for \textit{Association Phonétique Internationale}.}

Originally, the IPA alphabet was a list of symbols with pronunciation examples
from words in different languages. In 1900 the symbols were first organized into
a chart and were given phonetic feature labels (e.g.\ for manner of consonant
articulation among others `plosives', `nasales', `fricatives', for place of
articulation among others `bronchiales', `laryngales', `labiales' and for vowels
e.g.\ `fermées', `mi-fermées', `mi-ouvertes', `ouvertes'). Throughout the last
century, the structure of the chart has changed with increases in phonetic
knowledge. Thus, similar to notational systems in other scientific disciplines,
the IPA reflects facts and theories of phonetic knowledge that have developed
over time. It is natural then that the IPA is modified occasionally to
accommodate scientific innovations and discoveries. In fact, this is part of the
Association's mandate. These changes are captured in the recurring revisions to
the IPA.
%\footnote{For a detailed history see:
%\url{https://en.wikipedia.org/wiki/History\_of\_the\_International\_Phonetic_Alphabet}}

Over the years there have been several revisions, many minor. For example, the
articulation labels -- what often are call `features' even though the IPA
deliberately avoids this term -- have changed (e.g.\ terms like `lips', `throat'
or `rolled' are not used anymore). Phonetic symbol values have changed (e.g.\
voiceless is no longer marked by <h>). Symbols have been dropped (e.g.\ the
caret diacritic denoting `long and narrow' is no longer used). And many symbols
have been added to reflect contrastive sounds found in the world's very diverse
phonological systems. The current construction and use of the IPA are guided by
principles outlined in the \textit{Handbook of the International Phonetic
Association: A guide to the use of the International Phonetic Alphabet}
\citep{IPA1999}, henceforth simply called \textit{Handbook}.

@@

	\item it is intended to be a set of symbols for representing all possible
       sounds in the world's (spoken) languages;
	\item its chart has a linguistic basis (specifically a phonological bias)
       rather than a general phonetic notation scheme;
	\item its symbols can be used to represent distinctive feature
       combinations;\footnote{Although the chart uses traditional manner and
       place of articulation labels, the symbols can be used as a representation
       of any defined bundle of features, binary or otherwise, to define
       phonetic dimensions.}
	\item its chart provides a summary of linguists' agreed-upon phonetic 
	knowledge.

@@

Several styles of transcription with the IPA are possible, as illustrated in the
\textit{Handbook}, and they are all valid.\footnote{For illustration of
different possibilities, see the 29 languages and their transcriptions in the
original and initial \textit{Illustrations of the IPA}
\citep[41--154]{IPA2007}} Therefore, there are different but equivalent
transcriptions, or as \citet[64]{Ladefoged1990a} captures it, ``perhaps now that
the Association has been explicit in its eclectic approach, outsiders to the
Association will no longer speak of \textit{the} IPA transcription of a given
phenomenon, as if there were only one approved style.'' Clearly not all
phoneticians agree (or will likely ever agree of course) on all aspects of the
IPA or on transcription practices. As noted above, there have been several
revisions in the IPA's long history, but the current version (2005) is
strikingly similar to the 1926 version, showing the viability of the IPA as a
common standard for linguistic transcription.

@@

Any IPA transcription is based on two premises: (i) that it is possible to
describe the acoustic speech signal (sound waves) in terms of sequentially
ordered discrete segments, and, (ii) that each segment can be characterized by
an articulatory target.

Once spoken language data are segmented, the IPA provides symbols to
unambiguously represent phonetic details. However, since phonetic detail could
potentially include anything, e.g. something like `deep voice', the IPA
restricts phonetic detail to linguistically relevant aspects of speech.
Phonological consideration thus inextricably play a roll in transcription. In
other words, phonetic observations beyond quantitative acoustic analysis are
always made in terms of some phonological framework.

Today, the IPA chart reflects a linguistic theory grounded in principles of
phonological contrast and in knowledge about the attested linguistic variation.
This fact is stated explicitly in several places, including in the
\textit{Report on the 1989 Kiel convention} published in the \textit{Journal of
the International Phonetic Association} \citep[67-68]{International1989report}:

@@

From its earliest days the Association aimed to provide ``a separate sign for
each distinctive sound; that is, for each sound which, being used instead of
another, in the same language, can change the meaning of a word''
\citep[27]{IPA1999}. Distinctive sounds became later known as \textsc{phonemes}
and the IPA has developed historically into a notational devise with a strictly
segmental phonemic view. A phoneme is an abstract theoretical notion derived
from an acoustic signal as produced by speakers in the real world. Therefore the
IPA contains a number of theoretical assumptions about speech and how to
transcribe speech in written form.

@@

First, the symbols of IPA crucially distinguish between letter-like symbols and
diacritics. The use of the letter-like symbols to represent a language's sounds
is guided by the principle of contrast; where two words are distinguishable by
phonemic contrast, those contrasts should be transcribed with different letter
symbols (not diacritics). Allophonic distinction falls under the rubric of
diacritically-distinguished symbols, e.g. the difference in English between and
unrealized /p/ in [stop̚] vs. an aspirated /p/ in [spʰot] should be marked with
a diacritic. In sum:

@@

There are different reasons for this difficulty to interpret IPA symbols, all
officially sanctioned by the IPA. An important principle of the IPA is that
different representations resulting from different underlying analyses are
allowed. Because the IPA does not provide phonological analyses for specific
languages, the IPA does not define a single ``correct'' transcription system.
Rather, the IPA aims to provide a resource that allows users to express any
analysis so that it is widely understood.

Basically, the IPA allows for both a `narrow' phonetic transcription and a
`broad' phonological transcription. A `narrow' phonetic transcription may freely
use the all symbols in the IPA chart with direct reference to their phonetic
value, i.e. the transcriber can indicate with the symbols <ŋ͡m> that the
phonetic value of the attested sound is a simultaneous labial and velar closure
which is voiced and contains nasal airflow, independently of the phonemic status
of this sound in the language in question.
